Improving Haskell Theory Exploration

### 2015-11-23 TransferReport/background.tex
Added more feedback

\section{Background}
\label{background}

\subsection{Haskell}
\label{haskell}

@@

  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
  \label{coresyntax}
\end{figure}

@@

The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names. There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.

\subsection{QuickCheck}
\label{quickcheck}

Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g. life-) critical systems. To see why, consider the following Haskell function, along with some unit tests (see \ref{haskell} for more details on Haskell):

@@

We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}. These programs \emph{discover} properties of a ``theory'' (e.g. a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.\footnote{See section \ref{atp} for more information on automated theorem proving.}

\subsection{Theory Exploration}
\label{theoryexploration}

@@

  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_1$ to $N_n$ and types $T_1$ to $T_m$. We use $a =_{QC} b$ to denote that $a$ and $b$ are indistinguishable by \qcheck{}, and use $\ulcorner \dots \urcorner$ for quasiquotation.}
  \label{qspecmodel}
\end{figure}

@@

We focus on \qspec{} \citep{QuickSpec}, which discovers equations about Haskell code. These are found through the process modelled in figure \ref{qspecmodel}: given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.

@@

\subsection{Clustering and Feature Extraction}
\label{featureextraction}

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{relevance}. Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that theory exploration has no distinct \emph{goal} to compare expressions against. Instead, we are interested in relationships between \emph{any} components of a theory, and hence we must consider the relevance of \emph{everything} to \emph{everything else}. A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. See \S \ref{clustering} for more information.

@@

@@

In our case, we will use a novel feature extraction algorithm, described in \S \ref{recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.

>>>>

\section{Background}
\label{sec:background}

\subsection{Haskell}
\label{sec:haskell}

@@

  \caption{Simplified syntax of GHC Core in BNF style. $[]$ and $(,)$ denote repetition and grouping, respectively.}
  \label{fig:coresyntax}
\end{figure}

@@

The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names. There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.

\subsection{QuickCheck}
\label{sec:quickcheck}

Although unit testing is the de facto industry standard for quality assurance in non-critical systems, the level of confidence it provides is rather low, and totally inadequate for many (e.g. life-) critical systems. To see why, consider the following Haskell function, along with some unit tests (see \ref{sec:haskell} for more details on Haskell):

@@

We can reduce this bias further through the use of \emph{theory exploration} tools, such as \qspec{} and \hspec{}. These programs \emph{discover} properties of a ``theory'' (e.g. a library), through a combination of brute-force enumeration, random testing and (in the case of \hspec{}) automated theorem proving.\footnote{See section \ref{sec:atp} for more information on automated theorem proving.}

\subsection{Theory Exploration}
\label{sec:theoryexploration}

@@

  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_1$ to $N_n$ and types $T_1$ to $T_m$. We use $a =_{QC} b$ to denote that $a$ and $b$ are indistinguishable by \qcheck{}, and use $\ulcorner \dots \urcorner$ for quasiquotation.}
  \label{fig:qspecmodel}
\end{figure}

@@

We focus on \qspec{} \citep{QuickSpec}, which discovers equations about Haskell code. These are found through the process modelled in figure \ref{fig:qspecmodel}: \iffalse TODO: Figure just shows the syntax, it does not model any process \fi given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.

@@

\iffalse TODO: Not immediately clear how the problem of reducing the set of expressions relates to premise selection \fi
\iffalse TODO: This detailed comparison with premise selection is opaque without definiing/explaining what premise selection is \fi

\subsection{Clustering and Feature Extraction}
\label{sec:featureextraction}

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{sec:relevance}. Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that theory exploration has no distinct \emph{goal} to compare expressions against. Instead, we are interested in relationships between \emph{any} components of a theory, and hence we must consider the relevance of \emph{everything} to \emph{everything else}. \iffalse TODO: every term in the given signature? \fi A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. See \S \ref{sec:clustering} for more information.

@@

\iffalse TODO: Signatures or terms? \fi

\iffalse TODO: I would play up the clustering intelligence. Clustering is not just about breaking up, it is about discovering significant patterns in data. By forgetting this, you make motivation for your work sound too ``small'' \fi

@@

In our case, we will use a novel feature extraction algorithm, described in \S \ref{sec:recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.

\begin{figure}
  \centering
  \begin{tikzpicture}[scale=0.75, level/.style={sibling distance=60mm/#1}]
      % Tree
      \node {\texttt{Lam}}
        child {node {\texttt{"x"}}}
        child {node {\texttt{App}}
          child {node {\texttt{Var}}
            child {node {\texttt{Global}}
              child {node {\texttt{"base:Prelude.not"}}}}}
          child {node {\texttt{Var}}
            child {node {\texttt{Local}}
              child {node {\texttt{"x"}}}}}};

      % Arrow
      %\node[single arrow,
      %      draw=black,
      %      fill=black!10,
      %      minimum height=2cm,
      %      shape border rotate=0] at (0,-1) {};
      %\draw[-latex] (A.east) -- (B.west);

      % Feature vector

  \end{tikzpicture}

  \caption{}

  \label{fig:featureextractionpic}
\end{figure}

\iffalse
\subsection{Clustering}
% TODO
\fi

### 2015-11-23 TransferReport/implementation.tex
Updates
Added more feedback

\section{Implementation}

@@

\subsection{Circular Convolution}\label{circularconvolution}

@@

\subsection{\texttt{nix-eval}}\label{nixeval}

@@

\subsection{\textsc{ASTPlugin}}\label{astplugin}

@@

\begin{description}
  \item{Type variables}: In \S \ref{haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects. In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters. Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values. This difference from Haskell, which only has explicit binding of values, alters function properties like arity.

@@

  \item{Violating encapsulation}: As discussed in \S \ref{haskell}, Haskell allows names to be \emph{private} to a module. When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions. Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
\end{description}

@@

  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored. As mentioned in \S \ref{featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.

@@

  \item{Counterexample Generation}: As mentioned in \S \ref{tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing. Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g. as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing. As demonstrated in \S \ref{haskell}, a type's constructors might not be exported, in which case these methods fail. Also, the exponential complexity of such naive recursion makes these generators unusable in many cases. It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
\end{itemize}

\subsection{\textsc{MLSpec}}\label{mlspec}

@@

\subsection{\textsc{ML4HS}}\label{ml4hs}

@@

Inspired by the use of premise selection (see \S \ref{relevance}) to reduce the search space in ATP,
we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.

We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{clustering}), is an effective method for performing this separation.

>>>>

\section{Implementation}
\label{sec:implementation}

@@

\subsection{Circular Convolution}
\label{sec:circularconvolution}

@@

\subsection{\texttt{nix-eval}}
\label{sec:nixeval}

@@

\subsection{\textsc{ASTPlugin}}
\label{sec:astplugin}

@@

\begin{description}
  \item{Type variables}: In \S \ref{sec:haskell} we saw that parametric polymorphism can be thought of as values being parameterised by type-level objects. In System F, this is represented explicitly by a special binding form $\Lambda$, distinct from the $\lambda$ binding form for value parameters. Core also has explicit bindings for type-level objects, although for simplicity it uses the same $\lambda$ representation for both types and values. This difference from Haskell, which only has explicit binding of values, alters function properties like arity.

@@

  \item{Violating encapsulation}: As discussed in \S \ref{sec:haskell}, Haskell allows names to be \emph{private} to a module. When compiling Core, we have full access to private definitions, as well as references to private names from within other definitions. Hence the definitions we receive from \textsc{AstPlugin} will include private values which we cannot import into a theory exploration tool.
\end{description}

@@

  \item{Machine learning}: The Core s-expressions are used as input to our machine learning approaches, for guiding theory exploration using statistical properties of the theory being explored. As mentioned in \S \ref{sec:featureextraction}, it is inefficient to apply intensive machine learning algorithms directly to high-dimensional data representations, so we use various feature extraction algorithms to reduce their dimensionality.

@@

  \item{Counterexample Generation}: As mentioned in \S \ref{sec:tests}, there are many Haskell property checkers, which differ mostly in the way they generate values for testing. Since our theory exploration tools are built on these property checkers, we rely on the existence of value generators for at least some of the types in our theory; yet existing approaches to deriving such generators (e.g. as found in the \texttt{derive} package) rely on choosing an arbitrary constructor then recursing. As demonstrated in \S \ref{sec:haskell}, a type's constructors might not be exported, in which case these methods fail. Also, the exponential complexity of such naive recursion makes these generators unusable in many cases. It seems that a database of inspectable Haskell functions may be useful for solving this problem in a more practical way, although we have not yet performed experiments in this area.
\end{itemize}

\subsection{\textsc{MLSpec}}
\label{sec:mlspec}

@@

\subsection{\textsc{ML4HS}}
\label{sec:ml4hs}

@@

Inspired by the use of premise selection (see \S \ref{sec:relevance}) to reduce the search space in ATP,
we select sub-sets of the given theory to explore, chosen to try and keep together those expressions which combine in interesting ways, and to separate those which combine in uninteresting ways.

We hypothesise that similarity-based clustering of expressions, inspired by those of \textsc{ML4PG}  and related work in \textsc{ACL2(ml)} (see \S \ref{sec:clustering}), is an effective method for performing this separation.

### 2015-11-23 TransferReport/intro.tex
Added more feedback

As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity. The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis. For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).

@@

Functional programming is also well suited to more radical practices, such as \emph{property checking} and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet are currently prohibitively expensive. Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.

\iffalse
TODO
Our contributions are:

\begin{itemize}
  \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
  \item A
\end{itemize}
\fi

>>>>

\iffalse TODO: sounds strange... what does it mean for software to be capable or uncapable? of what? Rephrase? \fi As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity. The \emph{functional programming} paradigm has been proposed for addressing these issues \cite{hughes1989functional}, in part by constructing programs which are more amenable to mathematical analysis. \iffalse TODO: I am not sure the example makes sense without explaining what effects are and why separating them helps. Rather than going there, perhaps just say about declarative style of programming and types expressing prperties of computations? \fi For example, by separating \emph{computation} from \emph{effects}, we can analyse each in isolation; in particular, pure computations can be evaluated without fear of damaging side-effects, and results are repeatable (since there can be no dependence on external state).

@@

Functional programming is also well suited to less-widespread practices, such as \emph{property checking} (as popularised by \qcheck{}) and \emph{theorem proving}, which are promising methods for increasing confidence in software, yet can be prohibitively expensive. Here we investigate how the recent \emph{theory exploration} approach can lower the effort required to pursue these goals, and in particular how machine learning techniques can mitigate the costs of the combinatorial algorithms involved.

Our contributions are:

\begin{itemize}
  \item A framework for applying theory exploration tools such as \qspec{} to Haskell's existing package system.
  \item A feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf machine learning algorithms. \iffalse TODO: say how this item fits in the TE agenda. \fi
  \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
\end{itemize}

We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and the results of theory exploration systems like \qspec{} (\S \ref{sec:theoryexploration}). We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}). We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details \S \ref{sec:implementation}. A variety of related work is surveyed in \S \ref{sec:related}, and we give several potential directions for future research in \S \ref{sec:future} before concluding in \S \ref{sec:conclusion}.

### 2015-11-23 TransferReport/related.tex
Added more feedback

\section{Related Work}
\label{related}

\subsection{Haskell}
\label{haskelldesc}

@@

Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws. However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}). This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g. in the style of interactive theorem proving (see \S \ref{itp}), as these results may be invalidated by seemingly innocuous code changes. We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems. Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development. The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.

@@

\subsubsection{Artificial Curiosity}
\label{curiosity}

@@

  \label{wundt}
\end{figure}

@@

As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do. The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible. The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.

@@

\subsubsection{Relevance Filtering}
\label{relevance}

@@

\subsubsection{Clustering}
\label{clustering}

@@

\subsubsection{Interestingness in Concept Formation}
\label{conceptformation}

@@

\subsection{Interactive Theorem Proving}
\label{itp}

@@

\subsection{Automated Theorem Proving}
\label{atp}

@@

Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.

>>>>

\section{Related Work}
\label{sec:related}

\subsection{Haskell}
\label{sec:haskelldesc}

@@

Together, these features make Haskell code highly structured, amenable to logical analysis and subject to many algebraic laws. However, as mentioned with regards to type classes, Haskell itself is incapable of expressing or enforcing these laws (at least, without difficulty \citep{lindley2014hasochism}). This reduces the incentive to manually discover, state and prove theorems about Haskell code, e.g. in the style of interactive theorem proving (see \S \ref{sec:itp}), as these results may be invalidated by seemingly innocuous code changes. We will revisit the second-class status of theorem proving in Haskell in our discussion of theory exploration (\S \ref{sec:theoryexploration}), but we will remark that Haskell is thus in a unique position with regards to the discovery of interesting theorems. Namely that many discoveries may be available with very little work, simply because the code's authors are focused on \emph{software} development rather than \emph{proof} development. The same cannot be said, for example, of ITP systems; although our reasoning capabilities may be stronger in an ITP setting, much of the ``low hanging fruit'' will have already been found through the user's dedicated efforts, and hence TE would be more likely to fit into an automation role for the user, rather than providing truly unexpected new discoveries.

@@

\subsubsection{Artificial Curiosity}
\label{sec:curiosity}

@@

  \label{fig:wundt}
\end{figure}

@@

As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, which is very much what we would like a theory exploration system to do. The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible. The resulting behaviour is characterised by the \emph{Wundt curve} (shown in figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}.

@@

\subsubsection{Relevance Filtering}
\label{sec:relevance}

@@

\subsubsection{Clustering}
\label{sec:clustering}

@@

\subsubsection{Interestingness in Concept Formation}
\label{sec:conceptformation}

@@

\subsection{Interactive Theorem Proving}
\label{sec:itp}

@@

\subsection{Automated Theorem Proving}
\label{sec:atp}

@@

Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{sec:related}.

### 2015-11-23 TransferReport/report.tex
Added more feedback

\begin{abstract}
Theory Exploration is a promising approach to improving the quality and understanding of software, above that available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming. Current techniques are limited by their computational cost, which we mitigate through a divide and conquer approach: reducing the problem size whilst using a machine learning approach to preserve relevant details.
\end{abstract}

@@

\section{Future Work}
\label{future}

@@

\section{Conclusion}
\label{conclusion}

>>>>

\begin{abstract}
Theory Exploration is a promising approach to improving the quality and understanding of software, above that available through testing, in languages which are amenable to formal analysis such as those based on pure functional programming. Current techniques \iffalse TODO: it sounds as if you are talking about techniques of "formal analysis" here rather than techniques of TE. Reformulate. \fi are limited by their computational cost \iffalse TODO: Unclear what that means in this context. Computations of what is costly? why? same for "problem size" below \fi, which we mitigate through a divide and conquer approach: reducing the problem size whilst using a machine learning approach to preserve relevant details.
\end{abstract}

@@

\section{Future Work}
\label{sec:future}

@@

\section{Conclusion}
\label{sec:conclusion}
